Software variability in fMRI analysis:\\ comparing between-tool and numerical errors

### 2021-12-06 paper/paper.tex
More discussion

We verified the correctness of our analyses by comparing our unperturbed
t-statistic group maps with the ones obtained
in~\cite{bowring2019exploring}. For FSL, we found
identical checksums. For SPM, the checksums were different but differences
were visually unnoticeable. For AFNI, the activation maps were similar
overall, however, differences were more noticeable visually. 
The observed differences remained small (see Appendix ~\ref{sec:supp-repro}) and might be due to the use of
Octave vs MATLAB in SPM and of multithreading in AFNI. We performed visual quality control of the AFNI
and SPM results for each individual subject and confirmed that \tristan{Ali: which ones?} images were
correctly registered to the MNI template.

\subsection{In group analysis, BT variability was larger than and correlated with machine error}

Table~\ref{table:pipeline-stats} reports on summary statistics for the
group-level t-statistics. For each tool pair (A, B), BT variability was
significantly larger than machine error in tool A or B, in both thresholded
and unthresholded maps (Wilcoxon signed-rank test p~\textless~$10^{-162}$
for all tests, t-test p=0 for all tests). These global differences were
confirmed regionally (Figure~\ref{fig:unthresh-maps}-\textbf{A},\textbf{B},\textbf{C}). 
In addition, BT and WT variability appeared moderately correlated (R
in [0.56, 0.58], p=0.0, Figure~\ref{fig:unthresh-maps}-\textbf{D}). Interestingly, the correlation 
was driven by a set of voxels exhibiting comparable BT and
WT values (located around the identity line in the XY plot), which, however, did not seem to have 
any spatial consistency.

@@

  Table~\ref{table:pipeline-stats} also reports on summary statistics for subject-level unthresholded t-statistics maps. As for group-level
  maps, BT variability was consistently larger than WT variability (Wilcoxon
  signed-rank test p~\textless~$10^{-160}$ for all tests, t-test p=0 for all
  tests). However, for some subjects, machine error (WT variability) approached and even 
  surpassed BT in some regions (Figure~\ref{fig:unthresh-maps-sbj}).  As for the group maps, BT and WT variability appeared moderately
  correlated for all subjects (R in [0.442, 0.612], p=0.0 for all
  subjects, Appendix~\ref{sec:supp-subjects}).

@@

\subsection{At the precision t=17~bits, numerical variability approached BT variability in group analysis}

While the previous results were obtained for machine error, we also
evaluated numerical variability across different virtual precisions for FSL
and found that the virtual precision of t=17 bits minimized the RMSE
between BT and WT in unthresholded group maps. Figure~\ref{fig:gnp-mni} shows BT
and WT in the corresponding group maps obtained at t=17 bits, showing that
BT and WT reach comparable magnitudes in some regions. BT and WT remained
moderately correlated at this precision. \tristan{In the text, add statistics similar to Table 2.}

@@

  \subsection{Previous results are confirmed in thresholded group maps}

  Figure~\ref{fig:thresh-maps}-\textbf{A},\textbf{B},\textbf{C} compares BT
  and WT for thresholded group maps. Thresholding is an unstable operation
  that introduced high variability for both BT and WT at the edges of
  active regions. Except in these regions, BT remained consistently larger
  than WT. To measure a potential association between BT and WT, we
  measured WT instability and BT instability in each region of the
  HCP-MMP1.0 parcellation. The confusion matrices in
  Figure~\ref{fig:thresh-maps}-\textbf{D} report on such instabilities for
  the 360 tested regions. The average ratio of unstable regions was 26\%
  for BT and 9.2\% for WT, which confirmed that BT variability was larger than machine error.
  The average Cohen's kappa score between WT instability and BT instability was $\kappa=0.17$, indicating 
  a moderate agreement between WT instability and BT instability (-1 \textless $\kappa$ \textless 1, $\kappa$\leq 0 denotes chance agreement).

@@

In fMRI group analyses, machine error remains an order of magnitude
smaller than between-tool variability. Group analyses have a regularization
effect toward numerical noise, which is expected to amplify as sample size
increases. Therefore, for fMRI studies with large sample sizes, machine
error can safely be neglected with respect to between-tool variability. The
recommendation made in~\cite{botvinik2020variability} to rely on ``multiverse" analyses where
multiple analysis tools are compared is therefore likely to successfully correct
for machine error in group studies.

However, machine error remains important in group analyses that are based
on a single tool, as is commonly the case in current fMRI studies. In
particular, in our study, the inherent instability of thresholding was
triggered by machine error in 9\% of 360 brain regions, which indicates
that machine error might have impacted neuroscientific conclusions related
to these regions.

In subject analyses, machine error and between-tool variabilty can
become comparable for some subjects in some regions. This observation is particularly
relevant to the development of fMRI-based biomarkers aiming at
individualized phenotype predictions. Machine error is expected to play
a non-negligible role in such analyses, even when predictions combine
results produced by multiple tools.

The observed regularization effect of group analyses toward numerical noise 
is consistent with observations made in~\cite{kiar2020numerical} from diffusion MRI where 
connectome graph statistics were found to be substantially unstable at the subject-level 
while group distributions remained consistent.

\tristan{discuss larger numerical noise}

No association was found between
machine error and between-tool variability at the group or subject level,
which suggests that they can be investigated separately from each other.
Even though their effects can be of comparable magnitude, machine error and
between-tool variability are distinct and most likely involve different
root causes \tristan{revise this based on correlations}.

@@

\begin{figure*}[ht]
      \centering
      \includegraphics[width=.6\textwidth]{figures/sbj-std-corr-unthresh-plot.png}
      \caption{Relationship between BT and WT variabilty for 16 subjects.}
  \label{fig:unthresh-correlation-allsbj}
\end{figure*}

>>>>

We verified the correctness of our analyses by comparing our unperturbed
t-statistic group maps with the ones obtained
in~\cite{bowring2019exploring}. For FSL, we found
identical checksums. For SPM, the checksums were different but differences
were visually unnoticeable. For AFNI, the activation maps were similar
overall, however, differences were more noticeable visually. 
The observed differences remained small (see Appendix ~\ref{sec:supp-repro}) and might be due to the use of
Octave vs MATLAB in SPM and of multithreading in AFNI. We performed visual quality control of the AFNI
and SPM results for each individual subject and confirmed that T1-weighted images were
correctly skull-stripped and registered to the MNI template.

\subsection{In the group analysis, BT variability was larger than and correlated with machine error}

Table~\ref{table:pipeline-stats} reports on summary statistics for the
group-level t-statistics. For each tool pair (A, B), BT variability was
significantly larger than machine error in tool A or B, in both thresholded
and unthresholded maps (Wilcoxon signed-rank test p~\textless~$10^{-162}$
for all tests, t-test p=0 for all tests). These global differences were
confirmed across the maps (Figure~\ref{fig:unthresh-maps}-\textbf{A},\textbf{B},\textbf{C}). 
In addition, BT and WT variability appeared moderately correlated (Pearson's R
in [0.56, 0.58], p=0.0, Figure~\ref{fig:unthresh-maps}-\textbf{D}). Interestingly, the correlation 
was driven by a set of voxels exhibiting comparable BT and
WT values (located around the identity line in the XY plot), which, however, did not seem to have 
any spatial consistency \tristan{Ali, check for intensity differences between these voxels and the other ones}.

@@

  Table~\ref{table:pipeline-stats} also reports on summary statistics for
  subject-level unthresholded t-statistics maps. As for group-level maps,
  BT variability was consistently larger than WT variability (Wilcoxon
  signed-rank test p~\textless~$10^{-160}$ for all tests, t-test p=0 for
  all tests). However, Figure~\ref{fig:unthresh-maps-sbj} shows that for
  some subjects, machine error (WT variability) approached and even
  surpassed BT in some regions. As for the group maps, BT and WT
  variability appeared moderately correlated for all subjects (R in [0.442,
  0.612], p=0.0 for all subjects, see Appendix~\ref{sec:supp-subjects}).

@@

\subsection{At precision t=17~bits, WT variability approached BT variability in the group analysis}

While the previous results were obtained for machine error, we also
evaluated numerical variability across different virtual precisions for FSL
and found that the virtual precision of t=17 bits minimized the RMSE
between BT and WT in unthresholded group maps. Figure~\ref{fig:gnp-mni} shows BT
and WT in the corresponding group maps obtained at t=17 bits, showing that
BT and WT reach comparable magnitudes in some regions. BT and WT remained
moderately correlated at this precision. \tristan{In the text, add statistics similar to Table 2: is BT still larger than WT here?}

@@

  \subsection{Previous results were confirmed in thresholded group maps}

  Figure~\ref{fig:thresh-maps}-\textbf{A},\textbf{B},\textbf{C} compares BT
  and WT for thresholded group maps. Thresholding is an unstable operation
  that introduced variability for both BT and WT at the edges of
  active regions. Except in these regions, BT remained consistently larger
  than WT. To measure correlation between BT and WT, we
  measured WT instability and BT instability in each region of the
  HCP-MMP1.0 parcellation. The confusion matrices in
  Figure~\ref{fig:thresh-maps}-\textbf{D} report on such instabilities for
  the 360 tested regions. The average ratio of unstable regions was 26\%
  for BT and 9.2\% for WT, which confirmed that BT variability was larger than machine error.
  The average Cohen's kappa score\footnote{$\kappa \leq 0$ denotes chance agreement, -1 \textless $\kappa$ \textless 1} between WT instability and BT instability was $\kappa=0.17$, indicating 
  a moderate agreement between WT instability and BT instability.

@@

In fMRI group analyses, machine error remains an order of magnitude smaller
than between-tool variability. Group analyses have a regularization effect
toward numerical noise, which is expected to amplify as sample size
increases. Therefore, for fMRI studies with large sample sizes, machine
error can safely be neglected with respect to between-tool variability. The
recommendation made in~\cite{botvinik2020variability} to rely on
``multiverse" analyses where multiple analysis tools are compared is
therefore likely to successfully correct for machine error in group
studies. Nevertheless, machine error remains substantial in group analyses
that are based on a single tool, as is commonly the case in current fMRI
studies. In particular, in our study, the inherent instability of
thresholding was triggered by machine error in 9\% of 360 brain regions,
which indicates that machine error might have impacted neuroscientific
conclusions related to these regions.

In subject analyses, machine error and between-tool variabilty can become
of comparable magnitude for some subjects in some regions. This observation
is particularly relevant to the development of fMRI-based biomarkers aiming
at individualized phenotype predictions. Machine error is expected to play
a non-negligible role in such analyses, even when predictions combine
results produced by multiple tools. The observed regularization effect of
group analyses toward numerical noise is consistent with observations made
in~\cite{kiar2020numerical} from diffusion MRI where connectome graph
statistics were found to be substantially unstable at the subject-level
while group distributions remained consistent.

For both group and subject analyses, between-tool variability and machine
error were found to be moderately correlated. Even though between-tool
variability and numerical variability are different in nature, this result
suggests that in some cases both sources of variability may have a common
cause that might be related to the conditioning of the BOLD signal
estimation problem in a specific dataset. Therefore, in some regions,
instabilities of similar magnitude may be triggered by small numerical
perturbations, model variations or implementation differences. For
instance, the analysis of datasets with high motion may be unstable both 
across tools and numerically.

Therefore, numerical stability may be suitable as a proxy to study
between-tool variability. This speculation might be of practical value to
address software variability at large given that numerical stability refers
to a consistent mathematical framework whereas between-tool variability
remains more empirical. Likewise, numerical and between-tool stability may
be interesting proxies for analysis accuracy, and perhaps useful in quality
control.

The finding that numerical variability approached tool variability at the
virtual precision of t=17~bits is interesting too. Indeed, while machine
error generally introduces differences in the order of 1 ulp --- or
t=53~bits for double-precision values and t=24~bits for single-precision
ones --- common scientific software dependencies introduce larger errors.
For instance, SciPy's 2D spline interpolation was recently found to be
precise up to t=10~bits~\cite{pytracer} and might therefore introduce
numerical perturbations leading to errors in the range of between-tool
variability \tristan{Yohan, please check that sentence}. Such high errors
might be triggered by updates in operating systems, Python, MATLAB, and
other software dependencies.

Our results are limited by the type of numerical noise introduced in the
analyses. Indeed, we only introduced noise in the outputs of elementary
mathematical functions while numerical noise could creep in any
floating-point operation. Our estimation of numerical variability should
therefore be considered a lower bound. Our estimation of tool varability is
also likely to be underestimated, having tested only 3 analytical pipelines
among the thousands available~\cite{carp2012plurality}.

@@

\begin{figure*}[ht]
      \centering
      \includegraphics[width=.6\textwidth]{figures/sbj-std-corr-unthresh-plot.png}
      \caption{Comparison between BT and WT variabilty for 16 subjects.}
  \label{fig:unthresh-correlation-allsbj}
\end{figure*}
